---
title: "R Notebook"
output: html_notebook
---

Script for lake polygons, not regulated for hydropower use.

################################

## Loading required packages

################################


```{r load packages}
# Data management
library(here) # creating code with relative paths (all within the project folder)
library(rio) # convenience package for data import into R
library(tidyverse) # collection of packages that are loaded automatically: ggplot2, dplyr, tidyr,readr,purrr,tibble,stringr,forcats
library(knitr)
library(readxl) # read and write excel files
library(writexl)

# Graphics
library(sciplot) # bargraphs and lineplots
library(RColorBrewer) # color palettes, especially for nice looking maps

# Spatial files
library(sf) # spatial package
library(mapview) # interactive maps
library(geojsonsf) # converting between geojson and sf
library(rgdal) # for readOGR etc

```

################################

## Source files and import data

################################

## 1) NVE data

### NVE data: Innsjø
```{r download NVE data: innsjø }
## Data from NVE: innsjødatabase
innsjo_sf <- geojsonsf::geojson_sf(here::here("data","source_data","NVE_innsjo","NVEData","Innsjo_Innsjo.geojson"))

# Change to projected coordinates
innsjo_sf_P <- sf::st_transform(innsjo_sf, 32633)

```

### Data from GBIF: NTNU University Museum
```{r GBIF download INH dataset}
## Data from GBIF: The museum dataset for Norway
# GBIF.org (24 November 2022) GBIF Occurrence Download  https://doi.org/10.15468/dl.ee5z42
download_url <- "https://api.gbif.org/v1/occurrence/download/request/0176466-220831081235567.zip"
tmpfile <- tempfile()
tmpdir <- tempdir()
options(timeout = 500)
download.file(download_url,tmpfile)
occurrences <- rio::import(unzip(tmpfile,files="occurrence.txt",exdir = tmpdir), encoding = "UTF-8") # on the fly unzip and import to R object 

# Save locally
save(occurrences, file = here::here("data","source_data","occurrences.rda"))

```

```{r}
# Load saved file
load(file =  here::here("data","source_data","occurrences.rda"))

## Create a dataframe where each row is one location (based upon lat/long), summarized data
locations <- occurrences %>% 
  dplyr::group_by(locality, decimalLatitude,decimalLongitude) %>%
   dplyr::summarize(
    datasetName = paste0(unique(datasetName), collapse = ", "),
    publisher = paste0(unique(publisher), collapse = ", "),
    datasetID = paste0(unique(datasetID), collapse = ", "),
    datasetKey = paste0(unique(datasetKey), collapse = ", "),
    N_occurrences = length(unique(occurrenceID)),
    N_samplingEvents = length(unique(eventID)),
    locationIDs = paste(unique(locationID), collapse=", "),
    N_taxa = length(unique(scientificName)), # could change to taxonKey?
    N_methods = length(unique(samplingProtocol)),
    methods = paste0(unique(samplingProtocol), collapse = ", "),
    N_yrs = length(unique(year)),
    years = paste0(unique(year), collapse = ", "),
    first_year = min(year),
    last_year = max(year),
    period_yrs = (last_year - first_year),
    N_months = length(unique(month)),
    months = paste0(unique(month), collapse = ", "),
    field_number = paste0(unique(fieldNumber), collapse = ", "),
    collectionCode = paste0(unique(collectionCode), collapse = ", "),
    associatedReferences = paste0(unique(associatedReferences), collapse = ", ")
    ) 

# Make dataframe
locations <- as.data.frame(locations) 

## Make location data-frame a spatial object 
loc_sf <- st_as_sf(locations, coords = c("decimalLongitude","decimalLatitude"), crs = 4326, remove = FALSE)
# Create a version with projected coordinates
loc_sf_P <- sf::st_transform(loc_sf, 32633)
```


### Data from GBIF: Multiple datasets

Initial download from GBIF via API. When using this dataset please use the following citation:
GBIF.org (14 December 2022) GBIF Occurrence Download https://doi.org/10.15468/dl.59hjre
DOI: https://doi.org/10.15468/dl.59hjre (may take some hours before being active)
Creation Date: 15:00:41 14 December 2022
Records included: 722752 records from 15 published datasets
Compressed data size: 150.3 MB
Download format: DWCA

```{r GBIF data download}
## Data from GBIF: The museum dataset for Norway
# GBIF.org (24 November 2022) GBIF Occurrence Download  https://doi.org/10.15468/dl.ee5z42
download_url <- "https://api.gbif.org/v1/occurrence/download/request/0216182-220831081235567.zip"
tmpfile <- tempfile()
tmpdir <- tempdir()
options(timeout = 500)
download.file(download_url,tmpfile)
occurrences_mDat <- rio::import(unzip(tmpfile,files="occurrence.txt",exdir = tmpdir), encoding = "UTF-8") # on the fly unzip and import to R object 

# Save locally
save(occurrences_mDat, file = here::here("data","source_data","occurrences_mDat.rda"))

```

Load occurrence dataset and create spatial object
```{r load GBIF file}
# Load saved file
load(file =  here::here("data","source_data","occurrences_mDat.rda"))

# Remove rows with missing coordinates
occurrences_mDat <- occurrences_mDat %>% drop_na(decimalLongitude)
  # from 722752 to 708289 observations

## Create a dataframe where each row is one location (based upon lat/long), summarized data
locations_mDat <- occurrences_mDat %>% 
  dplyr::group_by(locality, decimalLatitude,decimalLongitude) %>%
   dplyr::summarize(
    datasetName = paste0(unique(datasetName), collapse = ", "),
    publisher = paste0(unique(publisher), collapse = ", "),
    datasetID = paste0(unique(datasetID), collapse = ", "),
    datasetKey = paste0(unique(datasetKey), collapse = ", "),
    N_occurrences = length(unique(occurrenceID)),
    N_samplingEvents = length(unique(eventID)),
    locationIDs = paste(unique(locationID), collapse=", "),
    N_taxa = length(unique(scientificName)), # could change to taxonKey?
    N_methods = length(unique(samplingProtocol)),
    methods = paste0(unique(samplingProtocol), collapse = ", "),
    N_yrs = length(unique(year)),
    years = paste0(unique(year), collapse = ", "),
    first_year = min(year),
    last_year = max(year),
    period_yrs = (last_year - first_year),
    N_months = length(unique(month)),
    months = paste0(unique(month), collapse = ", "),
    field_number = paste0(unique(fieldNumber), collapse = ", "),
    collectionCode = paste0(unique(collectionCode), collapse = ", "),
    associatedReferences = paste0(unique(associatedReferences), collapse = ", ")
    ) 

# Make dataframe
locations_mDat <- as.data.frame(locations_mDat) 

## Make location data-frame a spatial object 
loc_mDat_sf <- st_as_sf(locations_mDat, coords = c("decimalLongitude","decimalLatitude"), crs = 4326, remove = FALSE)
# Create a version with projected coordinates
loc_mDat_sf_P <- sf::st_transform(loc_mDat_sf, 32633)

```

### Merge occurrences dataframes to one file
```{r}
# Merge occurrence datasets: INH and mDat
occurrences_merged <- rbind(occurrences,occurrences_mDat)

# Save locally
save(occurrences_merged, file = here::here("data","source_data","occurrences_merged.rda"))

```

### Filtering
```{r}
# Load saved file
load(file =  here::here("data","source_data","occurrences_merged.rda"))



```

